---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic
glimpse(df_titanic)
```

**Observations**:

-   (List all variables here)

    The variables are Class, sex, age, survived and n/number of them (double).

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
-   Are there any differences?
    -   Yes, according to the wiki article there is 2224 indivduals that survived. But the data set has 2201.
-   If yes, what might account for those differences?
    -   Generally for large scale events like this things can happen like how data was retrieved. Such as reported from tickets sold, bodies

        collected etc. But still the difference is only about 20 people which is less than 1/100th of the total population.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
survivors <- df_titanic %>%
  filter(Survived == "Yes") %>%
  group_by(Class, Sex) %>%
  summarize(count = sum(n))
ggplot(survivors, aes(x = Class, y = count, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(title = "Count of Survivors", 
       x = "Class", 
       y = "Count of Survivors")
```

**Observations**:

-   I notice that the highest amount of survivors go to Male crew members and lowest go to female crew.
-   In most class cases minus crew, there were more female than male survivors.
-   3rd class was more even for male to female survivor split amount

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop2 <-
df_titanic %>%
  filter(Survived == "Yes") %>%
  group_by(Class, Sex) %>% 
  mutate(
    Total = sum(n),
    Prop = n / Total
  )
df_prop2
ggplot(df_prop2, aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(title = "Proportion of Survivors", 
       x = "Class", 
       y = "Proportion of Survivors")
```

**Observations**:

-   Write your observations here.

-   Is there anything *fishy* going on in your plot?
    -   I'm noticing it shows that according to this graph it shows that it is close to or nearly all the 1st class female, and crew (either male or female) survived given the amount of them present. Which I think what is going on is the graph is pulling data specifically from the first instance of the higher combination of sex and class to format the graph. So female and male crew children do not exist resulting in 0 survivors for them but proportionally all male and female crew that survived were adults so theyre both 100s.

-   <div>

    ### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

    </div>

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop2 <-
df_titanic %>%
  group_by(Class, Sex, Age) %>% 
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  filter(Survived == "Yes") 
df_prop2
ggplot(df_prop2, aes(x = Age, y = Prop, fill = Sex)) +
  facet_grid(cols = vars(Class)) + 
  geom_col(position = "dodge") +
  labs(title = "Proportion of Survivors", 
       x = "Class", 
       y = "Proportion of Survivors")
ggplot(df_prop2, aes(x = Class, y = Prop, fill = Sex)) +
  facet_grid(cols = vars(Age)) + 
  geom_col(position = "dodge") +
  labs(title = "Proportion of Survivors", 
       x = "Class", 
       y = "Proportion of Survivors")
```

**Observations**:

-   (Write your observations here.)
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   (Your response here)
    -   My new plot is odd because it shows that the way I am calculating for the proprtions, of all the male 2nd class 100% survivors are children and 8.33% are adults. This makes no sense because my total should add up to 100% as for the way I am calculating proportion from the total n.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
